---
title: 'Reproducible Research: Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---


## Loading and preprocessing the data

1. Load the data
```{r}

#   load the data
data <- read.csv("activity.csv")
str(data)
```
2. Process/transform the data (if necessary) into a format suitable for your analysis

```{r}
#   reduce the data with steps = NA and transform the date column to Date class
NAvalue <- is.na(data$steps)
subdata <- data[!NAvalue,]
subdata$date <- as.Date(subdata$date, "%Y-%m-%d")
str(subdata)

```


## What is mean total number of steps taken per day?
1. Make a histogram of the total number of steps taken each day
```{r}

# find the total steps taken each day
sum <- sapply(split(subdata, subdata$date), function(x){ sum(x$steps)})

# graph the histogram using barplot
barplot(sum, main = "Total steps each day", xlab = "date", ylab = "total steps")

```

2. Calculate and report the mean and median total number of steps taken per day
```{r}
mean <- sapply(split(subdata, subdata$date), function(x){ mean(x$steps)})
mean

median <- sapply(split(subdata, subdata$date), function(x){ median(x$steps)})
median
```
## What is the average daily activity pattern?
1. Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r}

# find the mean steps taken for each interval
average <- sapply(split(subdata, subdata$interval), function(x){ mean(x$steps)})

# graph the time series plot
plot(average, type = "l", main = "average steps taken vs 5-min interval", 
     xlab = "5-min interval", ylab = "average steps taken")

```


2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
average[average == max(average)]

```



## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r}
# use summary to find out the NA value in data$steps
summary(data)

```
    So we know that there are 2304 rows with NAs.

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r}
# copy the dataset
newdata <- data

# fill in missing data for the copy dataset
for(i in 1:length(NAvalue)){
    if(NAvalue[i] == TRUE){
        # replace the NAs with the mean for that day if exits, else replace with 0 
        name <- newdata$date[i]
        if(!is.na(unname(mean[name]))){
            newdata$steps[i] <- unname(mean[name])
        } else newdata$steps[i] <- 0 
    } #end of if     
} # end of for

# check if the new dataset contains NAs
summary(newdata)

```


4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r}

# find the total steps taken each day
newsum <- sapply(split(newdata, newdata$date), function(x){ sum(x$steps)})

# graph the histogram using barplot
barplot(newsum, col = "red", main = "Total steps each day", xlab = "date", ylab = "total steps")
# # add the histogram of the priviouse total using transparent black for comparing
# par(new = TRUE)
# barplot(sum, col = rgb(0,0,0,alpha=0.3), names.arg = "")

# get the new mean for each day
newmean <- sapply(split(newdata, newdata$date), function(x){ mean(x$steps)})
newmean

# get the new median for each day
newmedian <- sapply(split(newdata, newdata$date), function(x){ median(x$steps)})
newmedian

```
    We can see that after replacing the NAs with the mean for that day, the total steps, mean and median for each day change. For the total steps and mean, these values may be increased or decreased; for the median, all values increases. 

## Are there differences in activity patterns between weekdays and weekends?
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r}

# make another new copy of the dataset(the one with cleaning the NAs)
new2data <- subdata

# find out the weekdays of that date
new2data <- cbind(new2data, weekday = weekdays(new2data$date, abbreviate = TRUE))
new2data$weekday <- as.character(new2data$weekday)

# create a vector to identify the date of weekend
is_weekend <- new2data$weekday == "Sat" | new2data$weekday == "Sun"

# transformation
for(i in 1:length(is_weekend)){
    if(is_weekend[i] == TRUE) new2data$weekday[i] <- "Weekend"
    else new2data$weekday[i] <- "Weekday"
} # end of for

new2data$weekday <- as.factor(new2data$weekday)
summary(new2data)


```


2. Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). The plot should look something like the following, which was creating using simulated data:
```{r}

# find the mean steps taken for each interval

weekday <- new2data[new2data$weekday == "Weekday",]
weekend <- new2data[new2data$weekday == "Weekend",]

newaverage1 <- sapply(split(weekday, weekday$interval), function(x){ mean(x$steps)})
newaverage2 <- sapply(split(weekend, weekend$interval), function(x){ mean(x$steps)})

newave <- cbind("Weekday" = newaverage1, "Weekend" = newaverage2)

# graph the time series plot
library(lattice)
par(mfrow = c(2,1))
plot(newave , type = "l"
     , main = "average steps taken vs 5-min interval", 
     xlab = "5-min interval", ylab = "average steps taken")

```
